Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-04 article-models-tt.tex
added tags to model constructor rules, they look easier to parse now
working on the compatibility rules for the constructor of model type theory

\subsection{The basic ingredients of the type theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The following two rules essentially make a start with saying that every term is functorial
in the apropriate sense (there will be more rules contributing to this vies):}
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{f}{\mfam{P}}}{\mfam{f}}}\\
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\subst{\mfam{f}}{{f}{\mtm{P}}}}{\mtm{f}}}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
\end{align*}
We will call the terms that we introduced here the model constructors.

@@

@@

>>>>

\subsection{The basic ingredients of the type theory of models}
We first introduce the basic ingredients of our abstract theory of models.
\begin{align*}
& \inference
  {}
  {\jctx{\mctx}}
  \tag{the canonical model}\\
& \inference
  {\jfam{\mctx}{P}}
  {\jterm{\mctx}{P}{i}}
  \tag{initiality of $\mctx$}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jfam{{\Gamma}{A}}{\mfam{A}}}
  \tag{families}\\
& \inference
  {\jctx{\Gamma}}
  {\jfam{{\Gamma}{\mfam{\Gamma}}}{\mtm{\Gamma}}}
  \tag{terms}
\intertext{The following two rules essentially make a start with saying that every term is functorial
in the apropriate sense (there will be more rules contributing to this vies):}
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{f}{\mfam{P}}}{\mfam{f}}}\\
& \inference
  {\jterm{{\Gamma}{A}}{P}{f}}
  {\jterm{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\subst{\mfam{f}}{{f}{\mtm{P}}}}{\mtm{f}}}
\intertext{The empty context of a model $A$ is simply a term of $A$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{\Gamma}{A}{\tfemp{A}}}
  \tag{empty context}\\
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm{{{\Gamma}{A}}{\mfam{A}}}{\subst{\tfemp{\mfam{A}}}{\mtm{A}}}{\tft{A}}}
\intertext{Extension is going to be a morphism from $\ctxext{A}{\mfam{A}}$ to
$A$ in context $\Gamma$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
  \tag{extension}
\intertext{The action of weakening on families
is a morphism from $\ctxwk{\mfam{A}}{\mfam{A}}$ to
$\mfam{\mfam{A}}$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom{{{\Gamma}{A}}{\mfam{A}}}{\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}}
  \tag{weakening}
\intertext{The action of weakening on terms
is a morphism from $\ctxwk{\mfam{A}}{\mtm{A}}$ to $\mtm{\mfam{\mfam{A}}}$ over
$\tfwk{A}^0$ in context $\ctxext{{\Gamma}{A}}{\mfam{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
    {{{\Gamma}{A}}{\mfam{A}}}
    {\ctxwk{\mfam{A}}{\mfam{A}}}
    {\mfam{\mfam{A}}}
    {\tfwk{A}^0}
    {\ctxwk{\mfam{A}}{\mtm{A}}}
    {\mtm{\mfam{\mfam{A}}}}
    {\tfwk{A}^1}}
  \tag{weakening}
\intertext{The action of substitution by a term on families
is going to be a morphism from $\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}$ to 
$\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$
in context $\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}}
  \tag{substitution}
\intertext{Likewise, the action of substitution by a term on terms of those
families is going to be a morphism from $\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}$
to $\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}$ over $\tfsubst{A}^0$ in context
$\ctxext{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}$:}
& \inference
  {\jfam{\Gamma}{A}}
  {\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\tfsubst{A}^0}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mtm{A}}}}
  {\tfsubst{A}^1}}
  \tag{substitution}
\intertext{The identity functions are coded by}
& \inference
  {\jfam{\Gamma}{A}}
  {\jterm
   {{{\Gamma}{A}}{\mfam{A}}}
   {\subst{{\idfunc[\mfam{A}]}{\tfwk{A}^0}}{\mtm{\mfam{\mfam{A}}}}}
   {\tfid{A}}}
   \tag{identity functions}
\end{align*}
We will call the terms that we introduced here the model constructors.

@@

The easiest set of compatibility rules comes with weakening. The compatibility
rules that deal with substitution are likely going to have to do with the
Yoneda lemma, which we should be able to implement at some point.

\subsubsection{Families of weakenings}
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}}
  { \jfameq
    {{{\Gamma}{A}}{\ctxwk{A}{B}}}
    {\mfam{\ctxwk{A}{B}}}
    {\ctxwk{A}{\mfam{B}}}}.
\end{equation*}

\subsubsection{Terms of weakenings}
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    \jfam{\Gamma}{B}}
  { \jfameq
    {{{{\Gamma}{A}}{\ctxwk{A}{B}}}{\mfam{\ctxwk{A}{B}}}}
    {\mtm{\ctxwk{A}{B}}}
    {\ctxwk{A}{\mtm{B}}}}
\end{equation*}

@@

\subsubsection{The empty family over the empty family}
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jtermeq{\Gamma}{A}{\subst{\tfemp{A}}{\tfemp{\mfam{A}}}}{\tfemp{A}}}
\end{equation*}

\subsubsection{Extension by the empty family}
To understand the following inference rule, recall that for $\jfam{\Gamma}{A}$
we have the judgmental equalities
\begin{equation*}
\subst{\tfemp{\mfam{A}}}{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}\jdeq
\subst{\tfemp{\mfam{A}}}{\ctxwk{\mfam{A}}{{A}{A}}}\jdeq
\ctxwk{A}{A}.
\end{equation*}
Therefore, we can postulate:
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jhomeq{\Gamma}{A}{A}{\subst{\tfemp{\mfam{A}}}{\tfext{A}}}{\idfunc[A]}}
\end{equation*}

\subsubsection{Extensions of the empty context}
To understand the following inference rule, recall that for $\jfam{\Gamma}{A}$
we have the judgmental equalities
\begin{align*}
\subst{\tfemp{A}}{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}
& \jdeq \subst{\tfemp{A}}{\ctxwk{\mfam{A}}{{A}{A}}}\\
& \jdeq \ctxwk{\subst{\tfemp{A}}{\mfam{A}}}{\subst{\tfemp{A}}{\ctxwk{A}{A}}}\\
& \jdeq \ctxwk{A}{\subst{\tfemp{A}}{\ctxwk{A}{A}}}\\
& \jdeq \ctxwk{A}{A}
\end{align*}
Therefore, we can postulate
\begin{equation*}
\inference
  {\jfam{\Gamma}{A}}
  {\jhomeq{\Gamma}{A}{A}{\subst{\tfemp{A}}{\tfext{A}}}{\idfunc[A]}}
\end{equation*}

\subsubsection{Weakening by the empty family}

\subsubsection{Weakenings of the empty family}

\subsubsection{Substitutions of the empty family}

\subsubsection{Substitution by the term of the empty family}

\subsubsection{Extensions of extensions}

\subsubsection{Extensions of weakenings}

\subsubsection{Weakenings by extensions}

\subsubsection{Substitutions of extensions}
